Toward nonlinear multigrid \\ for nonlinear and nonlocal variational inequalities

### 2023-02-09 talk/slides.tex
changes from trial run

\hfill \mbox{
\includegraphics[height=0.15\textheight]{images/mg-vcycle.png} \quad
\includegraphics[height=0.15\textheight]{images/mg-fcycle.png}
}

@@

\begin{itemize}
\item implemented using an FD discretization and PETSc\footnote{Portable Extensible Toolkit for Scientific computing \quad \href{https://petsc.org/release/}{\texttt{petsc.org}} \quad \includegraphics[height=3mm]{images/petsc.png}}
    \begin{itemize}
    \item[$\circ$] FAS multigrid is a nonlinear solver (SNES) type
    \item[$\circ$] implement in C/Fortran/python/Julia(?)
    \end{itemize}
\item multigrid solvers in PETSc are \emph{composed} from smoothers on each level, and a solver on the coarse mesh
\item FAS F-cycle run:
\end{itemize}
\begin{Verbatim}[xleftmargin=15mm,fontsize=\scriptsize]
./bratu -da_grid_x 5 -da_grid_y 5 -da_refine J \
    -snes_rtol 1.0e-12 \
    -snes_type fas \
    -snes_fas_type full \
    -fas_levels_snes_type ngs \
    -fas_levels_snes_ngs_sweeps 2 \
    -fas_levels_snes_ngs_max_it 1 \
    -fas_levels_snes_norm_schedule none \
    -fas_coarse_snes_type ngs \
    -fas_coarse_snes_max_it 1 \
    -fas_coarse_snes_ngs_sweeps 4 \
    -fas_coarse_snes_ngs_max_it 1 
\end{Verbatim}

@@

\begin{itemize}
\item let's not get stuck on textbook example problems!
\item multigrid for a real-world VI problem?
\item consider a viscous layer with surface elevation $s(x,y)$ flowing with velocity $\bu(x,y,z)$, driven by gravity, over fixed bed topography with elevation $b(x,y)$, in a \emph{climate} which adds or removes fluid at a signed rate $a(x,y)$ [$\text{m}\,\text{s}^{-1}$]
    \begin{itemize}
    \item[$\circ$] data $a,b$ defined on domain $\Omega \subset \RR^2$
    \end{itemize}
\item geophysical examples: \alert{glaciers and ice sheets}, sea ice, lakes
\end{itemize}

@@

    \begin{itemize}
    \item[$\circ$] surface velocity $\bu|_s$ is determined by fluid domain geometry, including $s$ itself
    \item[$\circ$] $\bn_s=\left<-\grad s,1\right>$ is upward surface normal
    \item[$\circ$] generally: $-\bu|_s \cdot \bn_s$ is a \emph{non-local} function of $s$
    \end{itemize}
\item<2> the inequality constraint $s \ge b$ \alert{generates a free boundary}

@@

\begin{frame}{CP form (glacier ice in a climate)}

\begin{itemize}
\item VI form on previous slide is too abstracted for clarity
\item the strong form of the same problem is a \alert{complementarity problem (CP) coupled to a Stokes problem}:
\begin{align*}
s - b &\ge 0 && \text{ in $\Omega$} \\
- \bu|_s \cdot \bn_s - a &\ge 0 && \text{''} \\
(s-b)(- \bu|_s \cdot \bn_s - a) &= 0 && \text{''} \\
- \nabla \cdot \left(2 \nu(D\bu)\, D\bu\right) + \nabla p - \rhoi \mathbf{g} &= \bzero && \text{in $\Lambda(s)$} \\
\nabla \cdot \bu &= 0 && \text{''} \\
\left(2 \nu(D\bu) D\bu - pI\right) \bn &= \bzero && \{z=s\} \subset \partial \Lambda(s) \\
\bu &= \bzero && \{z=b\} \subset \partial \Lambda(s)
\end{align*}
\item solve this for $s$ on $\Omega$, and simultaneously for $\bu,p$ on $\Lambda(s) = \{(x,y,z) : b(x,y) < z < s(x,y)\}$
\end{itemize}
\end{frame}

@@

\begin{itemize}
\item[]
    \begin{itemize}
    \item[$\circ$] for more on this problem see (Bueler, 2021)
    \end{itemize}
\end{itemize}

\begin{itemize}
\item {\color{FireBrick} what is needed for scalable multilevel solutions?}
    \begin{enumerate}
    \item<2>[1.] iterates must be admissible
    \item<2>[2.] global linearization of $\Phi(s)$ must be avoided
    \item<2>[3.] smoother cost must be comparable to one residual    
    \end{enumerate}
\end{itemize}
\end{frame}

@@

\begin{itemize}
\item VIs are nonlinear problems, even for linear operators like $-\nabla^2$
\item Newton-multigrid is straightforward in PETSc:
\begin{Verbatim}[xleftmargin=13mm,fontsize=\scriptsize]
./obstacle -da_grid_x 3 -da_grid_y 3 \
    -snes_type vinewtonrsls -ksp_type cg \
    -pc_type mg -da_refine J
\end{Verbatim}
    \begin{itemize}
    \item[$\circ$] linear solver applies to inactive variables
        \begin{itemize}
        \item[] \texttt{rsls} = reduced space line search
        \end{itemize}
    \item[$\circ$] Newton step equations solved by CG with GMG V-cycles
    \end{itemize}
\item issue: the outer Newton iteration must converge on the active set \alert{before} multigrid can provide effective preconditioning
    \begin{itemize}
    \item[$\circ$] grid-dependent (growing) Newton iterations
    \end{itemize}
\end{itemize}

@@

\vspace{2mm}
\begin{Verbatim}[xleftmargin=13mm,fontsize=\scriptsize]
./obstacle -da_grid_x 3 -da_grid_y 3 \
    -snes_type vinewtonrsls -ksp_type cg \
    -pc_type mg
\end{Verbatim}

\vspace{-4.9mm}
\hspace{39mm} {\scriptsize \color{FireBrick} \texttt{-snes\_grid\_sequence J}}

>>>>

\only<1>{
\hfill \includegraphics[height=0.15\textheight]{images/mg-vcycle.png}
}
\only<2>{\hfill \includegraphics[height=0.15\textheight]{images/mg-fcycle.png} \quad \begin{tabular}{l} F-cycle = \\ nested iteration \end{tabular}
}

@@

\begin{itemize}
\item implemented using an FD discretization and PETSc\footnote{Portable Extensible Toolkit for Scientific computing \quad \href{https://petsc.org/release/}{\texttt{petsc.org}} \quad \includegraphics[height=3mm]{images/petsc.png}}
    \begin{itemize}
    \item[$\circ$] FAS multigrid is a nonlinear solver (SNES) type
    \item[$\circ$] implement in C/Fortran/python
    \end{itemize}
\item multigrid solvers in PETSc are \emph{composed} from smoothers on each level, and a solver on the coarse mesh
\item FAS F-cycle run:
\end{itemize}
\begin{Verbatim}[xleftmargin=15mm,fontsize=\scriptsize]
./bratu -da_grid_x 5 -da_grid_y 5 -da_refine J \
    -snes_rtol 1.0e-12 \
    -snes_type fas \
    -snes_fas_type full \
    -fas_levels_snes_type ngs \
    -fas_levels_snes_ngs_sweeps 2 \
    -fas_levels_snes_ngs_max_it 1 \
    -fas_levels_snes_norm_schedule none \
    -fas_coarse_snes_type ngs \
    -fas_coarse_snes_max_it 1 \
    -fas_coarse_snes_ngs_sweeps 4 \
    -fas_coarse_snes_ngs_max_it 1 
\end{Verbatim}

@@

\begin{itemize}
\item let's not get stuck on textbook example problems!
\item multigrid for a real-world VI problem?
\item consider an incompressible, viscous layer with surface elevation $s(x,y)$, flowing with velocity $\bu(x,y,z)$, driven by gravity, over fixed bed topography with elevation $b(x,y)$, in a \emph{climate} which adds or removes fluid at a signed rate $a(x,y)$ [$\text{m}\,\text{s}^{-1}$]
    \begin{itemize}
    \item[$\circ$] data $a,b$ defined on domain $\Omega \subset \RR^2$
    \end{itemize}
\item geophysical examples: \alert{glaciers and ice sheets}, sea ice, lakes
\end{itemize}

@@

    \begin{itemize}
    \item[$\circ$] surface velocity $\bu|_s$ is determined by fluid domain geometry, including $s$ itself
    \item[$\circ$] $\bn_s=\left<-\grad s,1\right>$ is upward surface normal
    \item[$\circ$] generally: $-\bu|_s \cdot \bn_s$ is a \emph{non-local} function of $s$
    \end{itemize}
\item the inequality constraint $s \ge b$ \alert{generates a free boundary}

@@

\begin{frame}{CP form of viscous fluid layer in a climate}

\begin{itemize}
\item VI form on previous slide is too abstracted for clarity
\item the strong form of the same problem is a \alert{complementarity problem (CP) coupled to a Stokes problem}:
\begin{align*}
s - b &\ge 0 && \text{ in $\Omega \subset \RR^2$} \\
- \bu|_s \cdot \bn_s - a &\ge 0 && \text{''} \\
(s-b)(- \bu|_s \cdot \bn_s - a) &= 0 && \text{''} \\
- \nabla \cdot \left(2 \nu(D\bu)\, D\bu\right) + \nabla p - \rhoi \mathbf{g} &= \bzero && \text{in $\Lambda(s) \subset \RR^3$} \\
\nabla \cdot \bu &= 0 && \text{''} \\
\left(2 \nu(D\bu) D\bu - pI\right) \bn &= \bzero && \{z=s\} \subset \partial \Lambda(s) \\
\bu &= \bzero && \{z=b\} \subset \partial \Lambda(s)
\end{align*}
\item solve this for $s$ on $\Omega$, and simultaneously for $\bu,p$ on $\Lambda(s) = \{b < z < s\}$
\end{itemize}
\end{frame}

@@

\begin{itemize}
\item[]
    \begin{itemize}
    \item[$\circ$] for more on this problem class see (Bueler, 2021)
    \end{itemize}
\end{itemize}

\begin{itemize}
\item {\color{FireBrick} what is needed for scalable multilevel solutions?}
    \begin{enumerate}
    \item[1.] iterates must be admissible
    \item[2.] global linearization of $\Phi(s)$ must be avoided
    \item[3.] smoother cost must be comparable to one residual    
    \end{enumerate}
\end{itemize}
\end{frame}

@@

\begin{itemize}
\item VIs are nonlinear problems, even for linear operators like $-\nabla^2$
\item Newton-multigrid is straightforward in PETSc:
\begin{Verbatim}[xleftmargin=13mm,fontsize=\scriptsize]
./obstacle -da_grid_x 3 -da_grid_y 3 \
    -snes_type vinewtonrsls -ksp_type cg -pc_type mg \
    -da_refine J
\end{Verbatim}
    \begin{itemize}
    \item[$\circ$] linear solver applies to inactive variables
        \begin{itemize}
        \item[] \texttt{rsls} = reduced space line search
        \end{itemize}
    \item[$\circ$] Newton step equations solved by CG with GMG V-cycles
    \end{itemize}
\item issue: the outer Newton iteration must converge on the active set \alert{before} multigrid can provide effective preconditioning
    \begin{itemize}
    \item[$\circ$] grid-dependent (growing) Newton iterations
    \end{itemize}
\end{itemize}

@@

\vspace{2mm}
\begin{Verbatim}[xleftmargin=13mm,fontsize=\scriptsize]
./obstacle -da_grid_x 3 -da_grid_y 3 \
    -snes_type vinewtonrsls -ksp_type cg -pc_type mg \
\end{Verbatim}

\vspace{-1.5mm}
\hspace{18.7mm} {\scriptsize \color{FireBrick} \texttt{-snes\_grid\_sequence J}}
